Validation of marker gene expression

### 2025-09-15 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
define cell line projects

```{r, results='asis'}
# check for if the project is a cell line project or not 
# explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
is_cell_line <- all(consensus_df$sample_type == "cell line") | params$project_id == "SCPCP000024"

@@

```{r, eval=!is_cell_line}
# table with one row per unique broad cell type/ marker gene combination 
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment 
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- consensus_df |> 
  # for each assigned cell type/marker gene combo get total detected and mean expression
  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
  dplyr::summarize(
    .by = c("broad_celltype_group", "ensembl_gene_id", "validation_group_annotation"),
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
  # now get the mean expression/ mean percentage across all marker genes for a given validation group
  # here the broad_celltype_group is the final assigned annotation for that group of cells 
  # the validation_group_annotation refers to the cell type that marker gene is associated with 
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::mutate(
      # get total percent expressed
      percent_exp = (detected_count / total_cells) * 100,
      all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100,
      # account for NA/unknowns and set axes order
      broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |>
        factor(levels = c(unique(markers_df$validation_group_annotation), "unknown"))
    )

>>>>

```{r, results='asis'}
# check for if the project is a cell line project or not 
is_cell_line <- params$project_id %in% c("SCPCP000020", "SCPCP000024")

@@

```{r, eval=!is_cell_line}
# table with one row per unique broad cell type/ marker gene combination 
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment 
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- consensus_df |> 
  # for each assigned cell type/marker gene combo get total detected and mean expression
  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
  dplyr::summarize(
    .by = c("broad_celltype_group", "ensembl_gene_id", "validation_group_annotation"),
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
  # now get the mean expression/ mean percentage across all marker genes for a given validation group
  # here the broad_celltype_group is the final assigned annotation for that group of cells 
  # the validation_group_annotation refers to the cell type that marker gene is associated with 
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::mutate(
      # get total percent expressed
      percent_exp = (detected_count / total_cells) * 100,
      all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100,
      # account for NA/unknowns and set axes order
      broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |>
        factor(levels = c(unique(markers_df$validation_group_annotation), "unknown"))
    ) |> 
  # remove any marker genes that don't have an assigned validation group
  tidyr::drop_na()
